Journey: A Study of Scalabilty of Memory Management in Linux (TODO: should we get a new title?)

### 2017-04-28 paper.tex
Minor edits

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space. All the physical pages in the system are
managed using the \texttt{struct page} descriptors in the kernel.  We do not
measure the overhead of \texttt{struct page} because it is known that there is
one struct for each physical page. We do not measure the overhead of page
tables because this can be calculated from the size of the virtual memory space
used by the process.  When a program makes a memory management system calls
such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
memory management requires measuring the overhead of these structs.

@@

Figure \ref{fig:mmap_latency} shows for various microbenchmarks, the time taken
per \texttt{mmap} operation as it varies with the total amount memory allocated
so far by the system.

@@

The frag benchmark takes the most time of any benchmark. When a region of the
address space is being expanded, the kernel is forced to unmap it and map a new
area elsewhere in the address space. This operation needs creation of a new
\texttt{vm\_area\_struct}.  The kernel also must search through the red-black
tree, though it never finds a node which can be expanded to include the new
mapping. Then, an insertion on the tree is needed, possibly including
rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} might need to be
deleted and removed from the tree. All of these operations ensure that frag
mmap pattern takes the most amount of time.

The strided workload is very similar to frag because it will also force the
kernel to always create a new structure and insert it into the tree after
traversing the tree first. However, it does not need to delete an existing
structure, because of which, it is faster than the frag benchmark.

Finally, the rand workload will also search through the red-black tree first,
but there is a reasonably high probability that it will find an existing
mapping adjacent to the desired mapping which can be expanded, obviating the
need to create a new struct and insert it into red-black tree. Therefore, most
of the times the time taken by rand benchmark is lower than strided.

@@

The cont workloads write to adjacent 4KB pages in the address space. A huge page
of size 2MB is allocated when a new page is touched initially. Subsequent 4KB
pages that are touched are subpages of the 2MB page allocated initially. As a
result, the memory used by the workload is almost equal to the amount of memory
it has written to. The RSS thus increases linearly until the system begins to
swap when the number of free pages falls below a threshold.

>>>>

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space. All the physical pages in the system are
managed using the \texttt{struct page} descriptors in the kernel.  We do not
measure the overhead of \texttt{struct page} because it is known that there is
one struct for each physical page. We do not measure the overhead of page
tables because this can be calculated from the size of the virtual memory space
used by the process.  When a program makes a memory management system call
such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
memory management requires measuring the overhead of these structs.

@@

Figure \ref{fig:mmap_latency} shows for various microbenchmarks the time taken
per \texttt{mmap} operation as it varies with the total amount memory allocated
so far by the workload.

@@

The frag benchmark takes the most time of any benchmark. When a region of the
address space is being expanded, the kernel is forced to unmap it and map a new
area elsewhere in the address space. This operation needs creation of a new
\texttt{vm\_area\_struct}.  The kernel also must search through the red-black
tree, though it never finds a node which can be expanded to include the new
mapping. Then, an insertion on the tree is needed, possibly including
rebalancing. Moreover, the existing \texttt{ vm\_area\_struct} needs to be
removed from the tree. All of these operations ensure that the frag workload
takes the most amount of time.

The strided workload is very similar to frag because it will also force the
kernel to always create a new structure and insert it into the tree after
traversing the tree first. However, it does not need to delete an existing
structure, which causes it to be faster than the frag benchmark.

Finally, the rand workload will also search through the red-black tree first,
but there is a reasonably high probability that it will find an existing
mapping adjacent to the desired mapping which can be expanded, obviating the
need to create a new struct and insert it into red-black tree. Therefore, usually
the time taken by rand benchmark is lower than strided.

@@

The cont workload writes to adjacent 4KB pages in the address space. A huge page
of size 2MB is allocated when a new page is touched initially. Subsequent 4KB
pages that are touched are subpages of the 2MB page allocated initially. As a
result, the memory used by the workload is almost equal to the amount of memory
it has written to. The RSS thus increases linearly until the system begins to
swap when the number of free pages falls below a threshold.
